---
title: "Tools to check for colorblind- and grayscale-friendliness"
author: "Kim Cressman"
date: "2026-03-05"
categories: [colorblind, grayscale, tools]
message: false
warning: false
draft: false
---

It's much easier to make colorblind-friendly graphs than it used to be; many thoughtful people have developed colorblind-friendly palettes. The site I most frequently see recommended for finding such a palette is [Color Brewer](https://colorbrewer2.org/). Those palettes are easy to access in R thanks to the `{RColorBrewer}` package, and they are great! But they were designed for maps, and.... I make mostly scatterplots. Unfortunately, the light colors in some of the most popular palettes just don't show up well as points on a graph. I was introduced to the [`{khroma}`](https://packages.tesselle.org/khroma) package a few years back and it has since become my favorite for finding colorblind-friendly palettes that do work well in scatterplots.  

The other day, a coworker who was using one of those great `{khroma}` palettes checked to see if it would print well in grayscale, by copying it into PowerPoint and converting it. And ..... it didn't work well. We hadn't known it, but **colorblind friendly doesn't necessarily equal grayscale-friendly**. I remembered some tools in R that let you investigate how your colors look to those with color deficiencies, and this post is about some of those resources.  

## Key Links  

- [`{colorblindr}` package](https://github.com/clauswilke/colorblindr) - for any palette and graph you can make in R, use one command (`cvd_grid()`) to simultaneously see what that graph looks like for the three main forms of colorblindness and in grayscale. You do this in your R session and can investigate graphs you've actually made.   
- [Color Palette Finder](https://r-graph-gallery.com/color-palette-finder) on R Graph Gallery - for an existing palette (as collected in the `{paleteer}` package), see how it looks in various types of graph, for the three main forms of colorblindness and in grayscale. This is online and can help you find a palette that works for all your needs.  
- [`{colorblindcheck}` package](https://jakubnowosad.com/colorblindcheck/articles/intro-to-colorblindcheck.html) - for any palette, easily view what it looks like for the three main forms of colorblindness. This package does not include grayscale functionality.


## Example of the packages in action  

### Installation  

`{colorblindcheck}` and `{khroma}` are simple:  

```{r}
#| eval: false

install.packages("colorblindcheck")
install.packages("khroma")
```


`{colorblindr}` is not on CRAN, so needs to be installed from GitHub. It depends on `{cowplot}` (also not on CRAN) and `{colorspace}`. These installation instructions are lightly modified from the [`{colorblindr}` README](https://github.com/clauswilke/colorblindr). They note it is important to install `{cowplot}` and `{colorspace}` *before* installing `{colorblindr}`.    

```{r}
#| eval: false

remotes::install_github("wilkelab/cowplot")
install.packages("colorspace")
remotes::install_github("clauswilke/colorblindr")
```


### Make up some data and a graph  

I find myself frequently working with 6 or so groups in my data, so that's the number I'll use here. I like the `muted` palette from `{khroma}` and though you can call it straight in `{ggplot2}` with `khroma::scale_color_muted()`, I also want to investigate the palette with other packages and so pull it out into the vector `color_pal`. 

```{r}
library(ggplot2)
library(colorblindcheck)
library(colorblindr)

color_pal <- khroma::color("muted")(6)
```

```{r}
set.seed(2026)
dat <- data.frame(group = rep(c("A", "B", "C", "D", "E", "F"), times = 5),
                  y = rnorm(30))

p <- ggplot(dat,
            aes(group, y, col = group)) +
    geom_jitter(size = 3, alpha = 0.7) +
    scale_color_manual(values = color_pal) +
    theme_classic()
p
```


### colorblindcheck  

If you don't care about grayscale, the simplest way to do this is with `{colorblindcheck}`.  

```{r}
colorblindcheck::palette_plot(color_pal)
```

I had to use the `::` notation because it turns out `{colorblindr}` has a function with the same name, and it only shows you the palette - not the extras that simulate colorblindness.  

`{colorblindcheck}` also generates a numeric table if you want more detail on the specs. If you set the argument `plot = TRUE` inside this function, you get both the table and the plot we already generated. See the [package vignette](https://cran.r-project.org/web/packages/colorblindcheck/vignettes/intro-to-colorblindcheck.html) for more details on this table.    

```{r}
colorblindcheck::palette_check(color_pal)
```

### colorblindr  

This is also very simple and quite possibly my new favorite R function. It re-makes *my graph*, simulating the different varieties of colorblindness and showing what it looks like in grayscale ('desaturated') too! We can see groups C and E are difficult to distinguish from each other in grayscale, and groups B, D, and F look pretty similar too.      

```{r}
colorblindr::cvd_grid(p)
```


### Additional note  

Of course there's more to a graph than point color - in the case of the simulated data frame and these graphs, I could easily use point shape to differentiate the groups in addition to color. That will also make the graph more robust to the similarities between groups in the grayscale version.

```{r}
p2 <- ggplot(dat,
             aes(group, y, 
                 col = group, 
                 shape = group)) +
    geom_jitter(size = 3, stroke = 1, alpha = 0.7) +
    scale_color_manual(values = color_pal) +
    theme_classic()
p2
```

Using `stroke` inside a `geom_point()` or similar layer makes the outlines of non-filled shapes thicker. This trick was taught to me by the same person that was interested in grayscale.  

And let's see how it looks all the different ways with `{colorblindr}`.  

```{r}
cvd_grid(p2)
```

Yep, works great!  

### Wrap-up  

This post focused on tools used within R. But if you haven't picked a color palette or made graphs yet, check out the [Color Palette Finder](https://r-graph-gallery.com/color-palette-finder) on R Graph Gallery - it's another great resource. Not only does it help with colors, but it's introduced me to packages like [`{nationalparkcolors}`](https://github.com/katiejolly/nationalparkcolors) and [`{fishualize}`](https://nschiett.github.io/fishualize/index.html) (new favorite package name? Lots of new favorites for me this week).  

I hope this is helpful! Good coding!  